Broom: Safe Region-Based Memory Management for Dataflow Programs

### 2015-09-24 paper/broom.tex
Sec 3. First commit.
Added more type rules.
First commit with *all* type rules.
Bugfixes in rules.
Getting rid of stmts - first commit.
Complete type system with lambdas.
Bugfixes. Added "new Region" type rule.

\documentclass{sigplanconf}

@@

%Formatting Commands
%-------------------
\newcommand{\C}[1]{\code{#1}}
\renewcommand{\k}[1]{\mathtt{#1}} % Code font in math mode.
\newcommand{\code}[1]{\,{\tt #1}\,}
\newcommand{\conj}{\wedge}
\newcommand{\disj}{\vee}
\newcommand{\seqn}[1]{\overline {#1}}
\newcommand{\tuplee}[1]{\langle #1 \rangle}
\newcommand{\ALT}{~\mid~}
\newcommand{\lamb}[2]{\lambda #1.\,#2}
\newcommand{\spc}[0]{\quad}
\newcommand{\U}[0]{\cup}
\newcommand{\X}[0]{\times}
\newcommand{\xn}[0]{\cap}
\newcommand{\redsto}[0]{\longrightarrow}
\newcommand{\rredsto}[0]{\rightsquigarrow_r }
\newcommand{\evalsto}[0]{\longrightarrow }
\newcommand{\env}[0]{\Gamma}
\newcommand{\finmaparrow}[0]{\overset { Fin }{ \longrightarrow } }
\newcommand{\envwf}[1]{\vdash\,#1}
\newcommand{\tywf}[2]{#1\,\vdash\,#2}
\newcommand{\hastyp}[3]{#1\,\vdash\,#2\,:\,#3}
\newcommand{\hassort}[3]{#1\,\vdash\,#2\,::\,#3}
\newcommand{\subtyp}[3]{#1\,\vdash\,#2 <: #3}
\newcommand{\generalizes}[3]{#1\,\vdash\,#2 \succ #3}
\newcommand{\msfolsig}[0]{{\Lambda \equiv}}
%\newcommand{\msentails}[2]{#1\,\models_L\,#2}
\newcommand{\msentails}[2]{#1\,\models\,#2}
\newcommand{\thesemof}[1]{ \llbracket #1 \rrbracket}
\newcommand{\defeq}[0]{ \triangleq }
\newcommand{\tymap}[0]{\mathcal{F}}
\newcommand{\etaeq}[0]{\eta_{eq}}
\newcommand{\semof}[1]{\thesemof{#1}}
\newcommand{\mssemof}[1]{\thesemof{#1}}
\newcommand{\rwsto}[0]{=}
\newcommand{\rrwsto}[0]{\hookrightarrow}
\newcommand{\rwbindfwd}[0]{\gamma_{\Rightarrow}}
\newcommand{\rwbindbwd}[0]{\gamma_{\Leftarrow}}
\newcommand{\eqhlt}[1]{\underline{#1}}
\newcommand{\tfr}[0]{F_R}
\newcommand{\ground}[1]{\underline{#1}}
\newcommand{\semeq}[0]{\equiv}
\newcommand{\elabsto}[0]{\looparrowright }
\newcommand{\arr}[0]{$\rightarrow$}
\newcommand{\eef}[0]{$\Leftrightarrow$}
\newcommand{\imp}[0]{$\Rightarrow$}
\newcommand{\RNull}[0]{\emptyset}
\newcommand{\cty}[0]{ty}
\newcommand{\hsp}[0]{\hspace*{-0.1in}}
\newcommand{\hspa}[0]{\hspace*{-0.08in}}
% removable:
\newcommand{\subst}[2]{\lbrack #1/#2 \rbrack}
\newcommand{\rel}[1]{{R}_{\mathit{#1}}}
\newcommand{\absof}[1]{\left\| #1 \right\| }
\newcommand{\broom}{{\sc Broom}\xspace}
\newcommand{\name}{{\sc Broom}\xspace}
\newcommand{\naiad}{{\sc Naiad}\xspace}
\newcommand{\csolve}{{\sc Csolve}\xspace}
\newcommand{\GK}[1]{{\textcolor{red}{{GK: #1}}}}
\newcommand{\csharp}{C\#~}
\newcommand{\outlives}{\succeq}

@@

In this paper, we demonstrate that the general purpose garbage
collection scheme employed by C\#.Net runtime is not very effective
for dataflow programs insofar as the latency due to GC is intolerably
high in many practical examples. We propose a memory management scheme
that combines user-managed memory regions with automatic garbage
collection, and lets programmers exploit specific memory behaviours
exhibited by dataflow programs. To protect programmers from the
dangers of manual memory management, we supplement our memory
management scheme with a region type system that statically ensures
the memory safety of region-annotated C\# programs, even in presence
of higher-order stateful functions. Further, we describe a type
inference algorithm that infers principal region types for first-order
programs thereby saving programmers the need to write region
annotations on types. Our inference algorithm is also very effective,
albeit incomplete, for higher-order programs. As a testament to
efficacy of our approach, we demonstrate performance improvements over
a range of realistic dataflow programs that were rewritten to use
region-based manual memory management, but are nevertheless judged to
be safe by our type system.

@@

>>>>

\documentclass[pldi]{sigplanconf}

@@

%Formatting Commands
%-------------------
\newcommand{\C}[1]{\code{#1}}
\renewcommand{\k}[1]{\mathtt{#1}} % Code font in math mode.
\newcommand{\code}[1]{\,{\tt #1}\,}
\newcommand{\conj}{\wedge}
\newcommand{\disj}{\vee}
\newcommand{\seqn}[1]{\overline {#1}}
\newcommand{\tuplee}[1]{\langle #1 \rangle}
\newcommand{\ALT}{~\mid~}
\newcommand{\lamb}[2]{\lambda #1.\,#2}
\newcommand{\spc}[0]{\quad}
\newcommand{\U}[0]{\cup}
\newcommand{\X}[0]{\times}
\newcommand{\xn}[0]{\cap}
\newcommand{\redsto}[0]{\longrightarrow}
\newcommand{\rredsto}[0]{\rightsquigarrow_r }
\newcommand{\evalsto}[0]{\longrightarrow }
\newcommand{\env}[0]{\Gamma}
\newcommand{\finmaparrow}[0]{\overset { Fin }{ \longrightarrow } }
\newcommand{\envwf}[1]{\vdash\,#1}
\newcommand{\tywf}[2]{#1\,\vdash\,#2 \; \texttt{ok}}
\newcommand{\fgjtywf}[2]{#1\,\Vdash\,#2 \; \texttt{ok}}
\newcommand{\isvalid}[2]{#1\,\vdash\,#2}
\newcommand{\hastyp}[3]{#1\,\vdash\,#2:#3}
\newcommand{\hasfgjtyp}[3]{#1\,\Vdash^{}\,#2\,:\,#3}
\newcommand{\hassort}[3]{#1\,\vdash\,#2\,::\,#3}
\newcommand{\subtyp}[3]{#1\,\vdash\,#2 <: #3}
\newcommand{\fgjsubtyp}[3]{#1\,\Vdash\,#2 <: #3}
\newcommand{\generalizes}[3]{#1\,\vdash\,#2 \succ #3}
\newcommand{\msfolsig}[0]{{\Lambda \equiv}}
%\newcommand{\msentails}[2]{#1\,\models_L\,#2}
\newcommand{\msentails}[2]{#1\,\models\,#2}
\newcommand{\thesemof}[1]{ \llbracket #1 \rrbracket}
\newcommand{\defeq}[0]{ \triangleq }
\newcommand{\tymap}[0]{\mathcal{F}}
\newcommand{\etaeq}[0]{\eta_{eq}}
\newcommand{\semof}[1]{\thesemof{#1}}
\newcommand{\mssemof}[1]{\thesemof{#1}}
\newcommand{\rwsto}[0]{=}
\newcommand{\rrwsto}[0]{\hookrightarrow}
\newcommand{\rwbindfwd}[0]{\gamma_{\Rightarrow}}
\newcommand{\rwbindbwd}[0]{\gamma_{\Leftarrow}}
\newcommand{\eqhlt}[1]{\underline{#1}}
\newcommand{\tfr}[0]{F_R}
\newcommand{\ground}[1]{\underline{#1}}
\newcommand{\semeq}[0]{\equiv}
\newcommand{\elabsto}[0]{\looparrowright }
\newcommand{\arr}[0]{$\rightarrow$}
\newcommand{\eef}[0]{$\Leftrightarrow$}
\newcommand{\imp}[0]{$\Rightarrow$}
\newcommand{\RNull}[0]{\emptyset}
\newcommand{\cty}[0]{ty}
\newcommand{\hsp}[0]{\hspace*{-0.1in}}
\newcommand{\hspa}[0]{\hspace*{-0.08in}}
% removable:
\newcommand{\subst}[2]{\lbrack #1/#2 \rbrack}
\newcommand{\absof}[1]{\left\| #1 \right\| }
\newcommand{\broom}{{\sc Broom}\xspace}
\newcommand{\name}{{\sc Broom}\xspace}
\newcommand{\fbname}{{\sc Featherweight Broom}\xspace}
\newcommand{\FB}{{\sc FB}\xspace}
\newcommand{\naiad}{{\sc Naiad}\xspace}
\newcommand{\csolve}{{\sc Csolve}\xspace}
\newcommand{\GK}[1]{{\textcolor{red}{{GK: #1}}}}
\newcommand{\csharp}{C\#~}
\newcommand{\outlives}{\succeq}
\newcommand{\coloneqq}{::=}
\newcommand{\R}[1]{\textrm{#1}}
\newcommand{\N}[1]{{\normalfont #1}}
\newcommand{\M}[1]{\mathtt{#1}}
\newcommand{\ObjZ}{\C{Object}}
\newcommand{\RgnZ}{\C{Region}}
\newcommand{\thisZ}{\C{this}}
\newcommand{\superZ}{\C{super}}
\newcommand{\unitZ}{\C{unit}}
\newcommand{\FuncZ}{\C{Func}}
\newcommand{\unitval}{\C{()}}
\newcommand{\inang}[1]{\langle #1 \rangle}
\newcommand{\rhoalloc}{\rho^a}
\newcommand{\rhobar}{\bar{\rho}}
\newcommand{\rhoallocm}{\rho^a_m}
\newcommand{\rhobarm}{\bar{\rho_m}}
\newcommand{\ralloc}{\pi^a}
\newcommand{\rgn}{\pi}
\newcommand{\rbar}{\bar{\pi}}
\newcommand{\tbar}{\bar{T}}
\newcommand{\taubar}{\bar{\tau}}
\newcommand{\xbar}{\bar{x}}
\newcommand{\vbar}{\bar{v}}
\newcommand{\ebar}{\bar{e}}
\newcommand{\extends}{\triangleleft}
\newcommand{\rextends}{\blacktriangleleft}
\renewcommand{\bar}[1]{\overline{#1}}
\newcommand{\angR}{\inang{\rhoalloc\rhobar \,|\, \phi}}
\newcommand{\angT}{\inang{\tbar}}
\newcommand{\tyvar}{a}
\newcommand{\tyvarb}{b}
\newcommand{\angAlpha}{\inang{\bar{\tyvar} \extends \bar{\fgjN}}}
\newcommand{\fgjN}{K}
\newcommand{\fbN}{N}
\newcommand{\headerOf}[1]{\C{class}\; #1\angR\angAlpha \extends \fbN}
\newcommand{\substFn}{\mathcal{S}}
\newcommand{\mang}{\inang{\rhoalloc_m\bar{\rho_m} \,|\, \phi_m}}
\newcommand{\rhoset}{\Sigma}
\newcommand{\rhoenv}{\Delta}
\newcommand{\aenv}{\Theta}
\newcommand{\phicx}{\Phi}
\newcommand{\subtypcx}{\rhoset,\rhoenv,\aenv,\phicx}
\newcommand{\A}{{\mathcal{A}}}
\newcommand{\exptycx}[2]{\A,#1,#2}
\newcommand{\void}{\C{void}}
\newcommand{\functy}[3]{\FuncZ\inang{#1}\inang{#2,#3}}
\newcommand{\stmtsemcx}{\A,\env}
\newcommand{\stmtsemcxp}{\A',\env'}
\newcommand{\stmtsemcxpp}{\A'',\env''}
\newcommand{\sredsto}{\Rightarrow}
\newcommand{\stmtsem}[4]{#1 \, \vdash \, #2 \,/\, #3 \sredsto #4}
\newcommand{\okin}[2]{#1 \; \texttt{ok \; in} \; #2}
\newcommand{\lambdaexp}[3]{\lambda\inang{#1}(#2).#3}
\newcommand{\Lambdaexp}[2]{\Lambda\inang{#1}.#2}
\newcommand{\letexp}[3]{\C{let}\;#1\,=\,#2\;\C{in}\;#3}
\newcommand{\letregion}[2]{\C{letregion}\;#1\;\C{in}\;#2}
\newcommand{\open}[3]{\C{open}\;#1\;\C{withroot}\;#2\;\C{in}\;#3}
\newcommand{\openalloc}[3]{\C{openalloc}\;#1\;\C{withroot}\;#2\;\C{in}\;#3}
\newcommand{\unpackexp}[4]{\C{let}\;(#1,\,#2)\,=\,\C{unpack}\;#3\;\C{in}\;#4}

@@

In this paper, we demonstrate that the general purpose garbage
collection scheme employed by C\#.Net runtime is not very effective
for dataflow programs to the extent that the latency due to GC is
intolerably high in many practical examples. We propose a memory
management scheme that combines user-managed memory regions with
automatic garbage collection, and lets programmers exploit specific
memory behaviours exhibited by dataflow programs. To protect
programmers from the dangers of manual memory management, we
supplement our memory management scheme with a region type system that
statically ensures the memory safety of region-annotated C\# programs,
even in presence of higher-order stateful functions. Further, we
describe a type inference algorithm that infers principal region types
for first-order programs thereby saving programmers the need to write
region annotations on types. Our inference algorithm is also very
effective, albeit incomplete, for higher-order programs. As a
testament to efficacy of our approach, we demonstrate performance
improvements over a range of realistic dataflow programs that were
rewritten to use region-based manual memory management, but are
nevertheless judged to be safe by our type system.

@@

\input{type-system}
